\section{Continuity, Part I}

\subsection{Definition of continuity}

A function $f$ is \textbf{continuous} at $a$ when

\[\lim_{x\to a}f(x)=f(a)\]

Inlining the limits definition, $f$ is continuous at $a$ if for all
$\epsilon>0$ there exists $\delta>0$ such that $0<|x-a|<\delta$ implies
$|f(x)-f(a)|<\epsilon$.

\vs

We can simplify this definition slightly. Observe that in continuous
functions $f(a)$ exists, and at $x=a$ we get $f(x)-f(a)=0$. Thus we
can relax the constraint $0<|x-a|<\delta$ to $|x-a|<\delta$.

\vs

A function $f$ is \textbf{continuous on an interval} $(a, b)$ if it's
continuous at all $c\in(a,b)$\footnote{Closed intervals are a tiny bit
  harder, and I'm keeping them out for brevity.}.

\subsection{Recognizing continuous functions}
The following theorems allow us to tell at a glance that large classes
of functions are continuous (e.g. polynomials, rational functions,
etc.)

\subsubsection*{Five easy proofs}

\paragraph{Constants.} Let $f(x)=c$. Then $f$ is continuous at all $a$
because
\[\lim_{x\to a}f(x)=c=f(a)\]

\paragraph{Identity.} Let $f(x)=x$. Then $f$ is continuous at all $a$
because
\[\lim_{x\to a}f(x)=a=f(a)\]

\paragraph{Addition.} Let $f,g\in\R\to\R$ be continuous at $a$. Then
$f+g$ is continuous at $a$ because
\[\lim_{x\to a}(f+g)(x)=\lim_{x\to a}f(x)+\lim_{x\to a}g(x)=f(a)+g(a)=(f+g)(a)\]

\paragraph{Multiplication.} Let $f,g\in\R\to\R$ be continuous at $a$. Then
$f\cdot g$ is continuous at $a$ because
\[\lim_{x\to a}(fg)(x)=\lim_{x\to a}f(x)\cdot\lim_{x\to a}g(x)=f(a)\cdot g(a)=(fg)(a)\]

\paragraph{Reciprocal.} Let $g$ be continuous at $a$. Then $\frac{1}{g}$
is continuous at $a$ where $g(a)\neq 0$ because
\[\lim_{x\to a}\left(\frac{1}{g}\right)(x)=\frac{1}{\lim_{x\to a}g(x)}=\frac{1}{g(a)}=\left(\frac{1}{g}\right)(a)\]

\subsubsection*{Slightly harder proof: composition}

Let $f,g\in\R\to\R$. Let $g$ be continuous at $a$, and let $f$ be
continuous at $g(a)$. Then $f\circ g$ is continuous at $a$. Put
differently, we want to show
\[\lim_{x\to a}(f\circ g)(x)=(f\circ g)(a)\]

Unpacking the definitions, let $\epsilon>0$ be given. We want to show there
exists $\delta>0$ such that $|x-a|<\delta$ implies
\begin{align*}
    |(f\circ g)&(x)-(f\circ g)(a)|\\
    &=|f(g(x))-f(g(a))|<\epsilon
\end{align*}

By problem statement we have two continuities.

\vs

\textbf{First}, $f$ is continuous at $g(a)$, i.e.
$\lim_{X\to g(a)}f(X)=f(g(a))$. Thus there exists $\delta'>0$ such that
$|X-g(a)|<\delta'$ implies $|f(X)-f(g(a))|<\epsilon$.

\vs

\textbf{Second}, $g$ is continuous at $a$, i.e.
$\lim_{x\to a}g(x)=g(a)$. Thus there exists $\delta>0$ such that
$|x-a|<\delta$ implies $|g(x)-g(a)|<\epsilon$. Since we can make
$\epsilon$ be anything, we can set it to $\delta'$.

\vs

I.e. there exists $\delta>0$ such that $|x-a|<\delta$ implies
$|g(x)-g(a)|<\delta'$. Intuitively, $g(x)$ is close to $g(a)$. But by the
first continuity, any $X$ close to $g(a)$ implies
\[|f(X)-f(g(a))|<\epsilon\]

Thus $|f(g(x))-f(g(a))|<\epsilon$, as desired.

\subsection{Appendix: Stars over Babylon}
Stars over Babylon is a modification of the Dirichlet function (see
\ref{subsubsec:dirichlet}).

\vs

\textbf{TODO}: come back and study this.

\subsection{Problems}

\subsubsection*{Problem 1}
For which of the following functions $f$ is there a continuous function $F$ with domain $\R$ such that $F(x)=f(x)$ for all $x$ in the domain of $f$?

\subsubsection*{Solution}
(i) $f(x)=\frac{x^2-4}{x-2}$
\[\frac{x^2-4}{x-2}=\frac{(x-2)(x+2)}{(x-2)}=x+2\]
when $x\neq 2$. Thus let $F(x)=x+2$. It follows $F(x)=f(x)$ for all $x$ in the domain of $f$, as $2$ is not in the domain of $f$. Further, $F$ is continuous in $R$ as
\[\lim_{x\to a}x+2=a+2\]

(ii) $f(x)=\frac{|x|}{x}$

There is no such function $F$. Observe that if we define $F$ such that $F(x)=f(x)$ for all $x$ in the domain of $f$, the only point we have left to manipulate is $0$. It's easy to see that no definition of $F(0)$ will allow $F$ to have a limit at $0$, and thus $F$ cannot be continuous.

\subsubsection*{Problem 5}
For each number $a$, find a function which is continuous at $a$, but not at any other points.

\subsubsection*{Solution}
As per textbook, the following modification of the Dirichlet function
is continuous at $0$ but nowhere else:
\[
f(x) = 
\begin{cases} 
x & \text{for rational } x,\\
0 & \text{for irrational } x.
\end{cases}
\]

Shifting horizontally by $a$ creates a function continous at $a$ but
nowhere else:
\[
g(x) = 
\begin{cases} 
x-a & \text{for rational } x,\\
0 & \text{for irrational } x.
\end{cases}
\]


\subsubsection*{Problem 8}
Suppose that $f$ is continuous at $a$ and $f(a)=0$. Prove that if $\alpha\neq 0$, then $f+\alpha$ is nonzero in some open interval containing $a$.

\subsubsection*{Solution}
Since $f$ is continous at $a$, by continuity theorems (addition of
identity and constant functions) $f+\alpha$ is also continuous. Also
$(f+\alpha)(a)=\alpha$, and thus $\lim_{x\to a}(f+\alpha)(x)=\alpha$. We can set
$\epsilon=\alpha$, and by limit definition there exists an interval in which
$|(f+\alpha)(x)-\alpha|<\alpha$. In another words within the interval $f+\alpha$ is less
than $\alpha$ away from $\alpha$, and thus cannot be zero.



%%% Local Variables:
%%% TeX-master: "notes"
%%% End:
